cILR: Taxonomic Enrichment Analysis with Isometric Log-Ratios

### 2021-05-28 tax_agg.tex

\section*{Background}  
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host. Previous research has shown that changes in the composition of the microbiome have been associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers often relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} would yield taxonomic abundance tables that would be part of downstream statistical analyses where associations between an outcome/exposure and identified taxa would be found.

However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilized multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel B of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a mixture distribution of two normal components. Panel A of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

\subsection*{Evaluation}
\subsubsection*{Parametric Simulations}  
To address the performance of cILR under different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

@@

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the relative nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

@@

In addition to simulation analyses, we also evaluated our method using real data sets across both 16S rRNA gene sequencing and whole-genome sequencing. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false discovery rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the Genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the Github repository associated with Calagaro et al. \cite{matteocalgaro2020}.

@@

\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). Additionally, we also utilized a similar data set from Gevers et al. \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages < 17) from the RISK cohort enrolled in North America. Of the 671 samples obtained, 500 samples belong to patients with IBD.

@@

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank-sum test. We added a pseudocount of 1 to all zero entry values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank-sum test.

@@

Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the cILR method against a naive Wilcoxon rank-sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios. Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach. False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site. For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank-sum test at $\alpha$ of 0.05. For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples. Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate. However, this resulted in lower power for adjusted cILR methods.}
    \label{fig:ss_hypo}
\end{figure}

However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B. In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un adjusted ones. However, in higher correlation scenarios, the difference in power mus much more dramatic. At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only perfomring at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%. These results indicated that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR.

These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}. Here, we tested the enrichment of aerobic microbes at each sample using approaches similar to our parametric simulations. As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods. Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Figure \ref{fig:ss_hypo}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated. Figure \ref{fig:ss_auc} presents this result. We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank-sum test. Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

@@

\subsection*{Differential abundance analysis}
Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank-sum test) to perform differential abundance analysis. We compare the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\subsection*{Disease Prediction} 
We can conceive of cILR as a feature engineering step before model fit and prediction. Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}). Additionally, we also compared our results with just using the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.

@@

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than both GSVA and ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

@@

These results demonstrated that cILR generated scores are informative features in disease prediction tasks. Simulation results indicated that cILR methods perform much better than both ssGSEA and GSVA, but not as good as the standard CLR approach. Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.

@@

\subsection*{Significance testing with cILR}
We can use cILR to test for enrichment of taxa-sets at the sample level. Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for. Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted. As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:ss_hypo}). The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation. Conversely, the unadjusted cILR methods and the naive Wilcoxon rank-sum test has inflated type I error even at low correlation levels ($\rho = 0.2$). However, the trade-off for proper type I error control is lower power. Simulation results showed that power is sensitive to sparsity across all correlation levels.

>>>>

\section*{Background}  
The microbiome is the collection of microorganisms (bacteria, protozoa, archaea, fungi, and viruses) which co-exist with their human host. Previous research has shown that changes in the composition of the microbiome are associated with important health outcomes such as inflammatory bowel disease \cite{proctor2019}, type II diabetes \cite{sharma2019}, and obesity \cite{aoun2020}. To understand the central role of the microbiome in human health, researchers have relied on high-throughput sequencing methods, either by targeting a specific representative gene (i.e. amplicon sequencing) or by profiling all the genomic content of the sample (i.e. whole-genome shotgun sequencing) \cite{cho2012}. Processing raw sequencing data using a variety of bioinformatic pipelines \cite{callahan2016,truong2015} yields taxonomic abundance tables that can be used in downstream statistical analyses to identify associations between an outcome/exposure and identified taxa.

However, there exists unique challenges in the analysis of these data tables \cite{li2019,li2015}. First, like other sequencing-based datasets, microbiome count data is often high dimensional, where the number of detected taxa far exceeds the number of samples usually present. For predictive tasks, microbiome-specific penalized regression approaches have been developed to address this issue \cite{shi2016}, allowing for simultaneous model fitting and variable selection. For differential abundance tasks, researchers often utilize multiple hypothesis correction methods \cite{sankaran2014,benjamini1995} or omnibus tests \cite{chen2018} to address hypothesis testing burden.

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of figure \ref{fig:kurt_gof}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a mixture distribution of two normal components. Panel B of figure \ref{fig:kurt_gof} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on cILR scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

\subsection*{Evaluation}
\subsubsection*{Parametric Simulations}  
To address the performance of cILR for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

@@

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

@@

In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false discovery rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}.

@@

\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). Additionally, we also utilized a similar data set from Gevers et al. \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada. Of the 671 samples obtained, 500 samples belong to patients with IBD.

@@

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the cILR method with a naive Wilcoxon rank sum test. We added a pseudocount of 1 to all zero entry values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared cILR methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test.

@@

Panel A and B in figure \ref{fig:ss_hypo} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the cILR method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted cILR variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted cILR methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
    \caption{Type I error rate \textbf{(A)}, and power \textbf{(B)} across different parametric simulation scenarios. Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach. False-positive rate \textbf{(C)} and true positive rate \textbf{(D)} evaluation of similar methods on real 16S rRNA data from the oral microbiome of the gingival site. For \textbf{(A)} and \textbf{(B)}, enrichment of a specified set was tested at the sample level using cILR and the Wilcoxon rank sum test at $\alpha$ of 0.05. For \textbf{(C)} and \textbf{(D)}, the set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples. Adjusted cILR demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate. However, this resulted in lower power for adjusted cILR methods.}
    \label{fig:ss_hypo}
\end{figure}

However, the trade-off for good type I error control is demonstrably lower power, as shown in figure \ref{fig:ss_hypo}B. In situations where there is no inter-taxa correlation, cILR still outperforms the wilcoxon rank sum test, however adjusted versions of cILR did not perform as well as un-adjusted ones. However, in higher correlation scenarios, the difference in power is much more dramatic. At the highest effect size (fold change of 3) and correlation ($\rho = 0.5$), adjusted cILR was only performing at 50\% power, while unadjusted cILR and wilcoxon rank sum test were able to reach 80\%. These results indicate that both sparsity and inter-taxa correlation impacts power, with correlation having a much more dramatic impact especially for adjusted versions of cILR. Most importantly, cILR demonstrate higher power in all scenarios where type I error is properly controlled.

These observations were replicated when assessed on the semi-labeled gingival data set from the Human Microbiome Project as described in \nameref{Methods}. Here, we tested the enrichment of aerobic microbes for each sample using approaches similar to our parametric simulations. As expected in Figure \ref{fig:ss_hypo}C, the proportion of falsely rejected hypotheses was high in the naive Wilcoxon test and unadjusted cILR methods. Conversely, adjusted cILR controls for false positives adequately at the correct $\alpha$ level of 0.05. Power analysis (Figure \ref{fig:ss_hypo}D) showed similar patterns, where unadjusted cILR methods and the Wilcoxon test have a higher proportion of null hypotheses correctly rejected, however, these results are not useful to a practitioner as the number of falsely rejected hypotheses are also equally high.

To further assess the utility of cILR in classifying samples with enriched sets, we generated AUC scores for different cILR scores using true labels of whether a sample has an inflated set. This analysis, therefore, assessed the relative ranking of samples using cILR scores whereby high scores should correspond to samples that are known to be inflated. Figure \ref{fig:ss_auc} presents this result. We compared different variants of cILR against competing methods in the gene set testing space (GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}), as well as the $U$ test statistic from the Wilcoxon rank sum test. Across both simulations (Figure \ref{fig:ss_auc}A) and real-data applications (Figure \ref{fig:ss_auc}B), cILR scores perform marginally better especially in low effect size situations but did not stand out in most other scenarios. In simulation studies, classification performance was good (around AUC of 0.8) even at high correlation settings, only requiring medium effect sizes (fold change of 2). Notably, the W-statistic provided the least information for classifying samples with inflated taxa.

@@

\subsection*{Differential abundance analysis}
Here we assessed the ability to use cILR with a standard difference of means statistical test (such as Welch's t-test or a Wilcoxon rank sum test) to perform differential abundance analysis. We compared the performance of cILR related methods to two commonly used approaches in the microbiome literature: DESeq2 \cite{love2014} and corncob \cite{martin2020}.

\subsection*{Disease Prediction} 
We can conceive of cILR as a feature engineering step before model fitting and prediction. Here we assessed the predictive performance of a standard machine learning model (random forest \cite{breiman2001}) using cILR scores compared to scores generated by other single sample enrichment methods in the gene set testing literature (ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}). Additionally, we also compared our results with the use of the centered log-ratio (CLR) transformation on taxa sets aggregated via count summations.

@@

In addition to parametric simulations, we also demonstrated predictive performance using real data sets in Figure \ref{fig:pred_data}. The learning task across both data sets is to classify patients who are diagnosed with inflammatory bowel disease (includes both Crohn's disease and ulcerative colitis) and healthy controls. The Gevers et al. \cite{gevers2014} data set is a 16S rRNA sequencing data set while the Nielsen et al. \cite{nielsen2014} is a whole-genome shotgun sequencing data set. Similar to simulation experiments, across both data sets cILR methods provide much better performance than either GSVA or ssGSEA. Interestingly, the standard CLR approach only outperformed cILR in the 16S rRNA data set but was marginally worse in the WGS data set. Additionally, we observed that using the normal distribution for cILR generates better predictive performance compared to the other variants.

@@

These results demonstrate that cILR generated scores are informative features in disease prediction tasks. Simulation results indicate that cILR methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach. Interestingly, however, cILR methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.

@@

\subsection*{Inference with cILR}
We can use cILR to test for enrichment of taxa-sets at the sample level. Inter-taxa correlation can result in an inflated variance of resulting test statistics \cite{wu2012}, which results in an inflated type I error if not properly controlled for. Our method addresses this issue by combining the mean estimate of scores computed on permuted data and the variance estimate of scores computed on unpermuted data, where correlation is undisrupted. As a consequence, adjusted cILR methods control very well for type I error at the appropriate $\alpha$ level (Figure \ref{fig:ss_hypo}). The conservativeness of the test persisted across different sparsity levels, as well as inter-taxa correlation. Conversely, the unadjusted cILR methods and the naive Wilcoxon rank sum test has inflated type I error even at low correlation levels ($\rho = 0.2$). However, the trade-off for proper type I error control is lower power. Simulation results showed that power is sensitive to sparsity across all correlation levels.
